{\bf Stateless Model Checking \\ with Data-Race Preemption Points}

### 2016-03-01 oopsla/calcm.tex
Add beginnings of CALCM seminar talk

\title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
\subtitle[]{ {\em more clever than } ``\texttt{mandelbrot}'' {\em since 2011.}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[CMU 15-410]{Carnegie Mellon University - 15-410}
\date[]{2016, February 10}

@@

\section{Introduction}
\subsection{Introduction}

\begin{frame}{Outline}
	\textbf{Theory: Seeing race conditions in a new way}
	\begin{itemize}
		\item Case study (example)
		\item Tabular execution traces
		\item The execution tree
	\end{itemize}
	{\bf Research Technique: ``Systematic testing''}
	% Say: A way of enumerating all possible interleavings of a concurrent test case.
	\begin{itemize}
		\item Preemption points
		\item Challenges and feasibility
	\end{itemize}
	{\bf Tool: Landslide}
	\begin{itemize}
		\item How it works
		\item Automatically choosing preemption points
		\item User study (that's you!)
			% !!! SAY !!! : If you are struggling, this is not for you, etc.
	\end{itemize}
\end{frame}

\subsection{Race Conditions}

\definecolor{thread1}{RGB}{87,172,255}
\definecolor{thread2}{RGB}{255,201,102}
\definecolor{thread3}{RGB}{255,128,160}

@@

		\begin{itemize}
			\item See {\bf Paradise Lost} lecture!
			\item {\tt if} vs {\tt while}: Two consumers can race to make one fail.
		\end{itemize}
	\end{center}
\end{frame}
\begin{frame}{Thread Interleavings (``good'' case)}
	\begin{center}
		\begin{tabular}{|l|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
			\hline
			\small \texttt{lock(mx);} & & \\
			\small \texttt{if~(!work\_exists())} & & \\
			\small \texttt{~~~~wait(cvar,~mx);} & & \\

@@

\begin{frame}{Testing}
	How can programmers be confident in the correctness of their code?
	\begin{itemize}
		\item Unit tests
		\begin{itemize}
			\item good for basic functionality, bad for concurrency
		\end{itemize}
		\item Stress tests
		\begin{itemize}
			\item state of the art in 15-410
		\end{itemize}
		\item Theorem proving
		\begin{itemize}
			\item heavy burden on the programmers
		\end{itemize}
		\item Releasing to paying customers and worrying about correctness later
			% Say... "and I'm hoping to prevent you from finding yourself in a future where this happens!"
	\end{itemize}
	\linegap

	{\bf Motivation}: Can we do better than stress testing?
\end{frame}

\begin{frame}{Testing Mechanisms}
	\textbf{Stress testing}: \texttt{largetest}, \texttt{mandelbrot} and friends
	\begin{itemize}
		\item Attempting to exercise as many interleavings as practical
		\item Exposes race conditions at random
		\begin{itemize}
			\item ``If a preemption occurs at just the right time\ldots''
		\end{itemize}
		\item Cryptic panic messages when failure occurs
	\end{itemize}
	\linegap
	What if\ldots
	\begin{itemize}
		\item Make educated guesses about when to preempt
		\item Preempt enough times to run {\em every single} interleaving
		\item Tell the story of what {\em actually happened}.
		\item Overlook fewer bugs!
	\end{itemize}
	% Say: At the end of this lecture, I'm going to be looking for some among you...
\end{frame}

\section{Systematic Testing}

\breakslide{\Large A different way of looking at race conditions\ldots}

\begin{frame}{Execution Tree}
		\includegraphics[width=0.78\textwidth]{execution-tree-0.pdf} \\
		\vspace{-3in}
		\hspace{3in}
		\includegraphics[width=0.34\textwidth]{table-0.png}
		\\
		\vspace{1.2in}
		\hspace{3.7in}
		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
\end{frame}
\begin{frame}{Execution Tree}
		\includegraphics[width=0.78\textwidth]{execution-tree-1.pdf} \\
		\vspace{-3in}
		\hspace{3in}
		\includegraphics[width=0.34\textwidth]{table-0.png}
		\\
		\vspace{1.2in}
		\hspace{3.7in}
		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
\end{frame}
\begin{frame}{Execution Tree}
		\includegraphics[width=0.78\textwidth]{execution-tree-2.pdf} \\
		\vspace{-3in}
		\hspace{3in}
		\includegraphics[width=0.34\textwidth]{table-1.png}
		\\
		\vspace{1.2in}
		\hspace{3.7in}
		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
\end{frame}
\begin{frame}{Execution Tree}
		\includegraphics[width=0.78\textwidth]{execution-tree-full.pdf} \\
		\vspace{-3in}
		\hspace{3in}
		\includegraphics[width=0.34\textwidth]{table-2.png}
		\\
		\vspace{1.2in}
		\hspace{3.7in}
		\includegraphics[width=0.15\textwidth]{frownie.png}
\end{frame}

% Say: Obviously we want to be able to always run the third path. But knowing
% that ahead of time is impossible.

% Say: In the rest of the lecture I'm going to refer to the "state space of
% possible interleavings", simply referring to this tree, and "size" of the state
% space refers to the number of branches.

\begin{frame}{Systematic Testing - The Big Picture}
	\textbf{Goal: Force the system to execute every possible interleaving.}
	% Say: checking for crashes/memory errors/etc in each one.
	\begin{itemize}
		\item On 1st execution, schedule threads arbitrarily until program ends.
			\begin{itemize}
				\item This represents one branch of the tree.
			\end{itemize}
		\item At end of each branch, rewind system and restart test.
		\item Artificially add preemptions to produce different thread interleavings.
		\item Intuitively: Generate many ``tabular execution traces''.
	\end{itemize}
	\pause
	\linegap

	Okay, wait a sec...
	\begin{itemize}
		\item How can you possibly execute {\em every possible} interleaving?
		%\item How can you know where preempting matters most?
		\item How did you know to draw that tree's branches where they matter?
	\end{itemize}
\end{frame}

\begin{frame}{Preemption Point Example (remember this?)}
	\fbox{
	\begin{tabular}{l}
	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
	\\
	\begin{tabular}{rll}
		1 & \texttt{want[i]~=~true;} & \\
		  & & \footnotesize (preemption point A) \\
		2 & \texttt{while~(want[j])} & \\
		  & & \footnotesize (preemption point B) \\
		3 & \texttt{~~~~~~~~continue;} & \\
		  & & \footnotesize (preemption point C) \\
		4 & \texttt{//~...critical~section...} & \\
		  & & \footnotesize (preemption point D) \\
		5 & \texttt{want[i]~=~false;} & \\
	\end{tabular}
	\end{tabular}
	}
	\linegap

	{\bf Some preemption points will expose bugs. \\
	Some preemption points don't matter.}

\end{frame}

\begin{frame}{Preemption Point Example (remember this?)}
	\fbox{
	\begin{tabular}{l}
	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
	\\
	\begin{tabular}{rll}
		1 & \texttt{want[i]~=~true;} & \\
		  & & \footnotesize (preemption point A) \\
		2 & \texttt{while~(want[j])} & \\
		  & & \footnotesize ~ \\
		3 & \texttt{~~~~~~~~continue;} & \\
		  & & \footnotesize ~ \\
		4 & \texttt{//~...critical~section...} & \\
		  & & \footnotesize ~ \\
		5 & \texttt{want[i]~=~false;} & \\
	\end{tabular}
	\end{tabular}
	}
	\linegap

	Here, only preemption point A will trigger a deadlock. {\bf ~}\\
	All other interleavings are benign.{\bf ~}

\end{frame}

\begin{frame}{Preemption Points}
	\textbf{Preemption points} (PPs) are code locations where being preempted may cause different behaviour.
	\begin{itemize}
		\item IOW, somewhere that interesting interleavings can happen around.
	\end{itemize}
	\linegap

	Systematic tests are {\em parameterized} by the set of PPs.
	\begin{itemize}
		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
		\item Need to choose the set of PPs very carefully for test to be effective.
		\begin{itemize}
			\item ``Effective'' = both comprehensive and feasible.
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Preemption Points}
	What does ``all possible interleavings'' actually mean?

	\linegap
	{\bf One extreme}: Preempt at {\em every instruction}
	\begin{itemize}
		\item Good news: Will find every possible race condition.
		\item Bad news: Runtime of test will be impossibly large.
	\end{itemize}
	\linegap

	{\bf Other extreme}: {\em Nothing} is a preemption point
	\begin{itemize}
		\item Good news: Test will finish quickly.
		\item Bad news: Only one execution was checked for bugginess.
		\begin{itemize}
			\item No alternative interleavings explored.
			\item Makes ``no race found'' a weak claim.
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Preemption Points}
	\textbf{Sweet spot}: Insert a thread switch everywhere it ``might matter''.

	\linegap
	When do we fear being preempted?
	\begin{itemize}
		\item Threads becoming runnable (\texttt{thr\_create()}, \texttt{cond\_signal()}, etc.)
			\begin{itemize}
				\item Preemptions may cause it to run before we're ready
			\end{itemize}
		\item Synchronization primitives (\texttt{mutex\_lock()}/\texttt{unlock()}, etc.)
			\begin{itemize}
				\item If buggy or used improperly\ldots
			\end{itemize}
		\item Unprotected shared memory accesses (``data races'')% Say: more on this later.
			\begin{itemize}
				\item May result in data structure corruption
				\item More on this later...
			\end{itemize}
	\end{itemize}

	%%%% Not anymore! :D
	%Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
	%However, all these points together may still be too expensive...
\end{frame}

%\begin{frame}{Challenges}
%	Parameters of systematic tests must be kept small.
%	\begin{itemize}
%		\item Test program length / number of threads
%		\item Number of preemption points used
%	\end{itemize}
%	\linegap
%
%	\begin{center}
%		\hspace{-0.25in}\includegraphics[width=0.3\textwidth]{expo.pdf} \\
%		\vspace{-1.33in}\hspace{-1.75in}{\rotatebox{90}{\small state space size}} \\
%		\vspace{0.15in}\hspace{-0.25in}{\small number of PPs}
%	\end{center}
%\end{frame}

\section{Landslide}

\breakslide{\Huge Landslide}

\begin{frame}{About The Project}
	\textbf{About me}: 5th year graduate student, advised by Garth Gibson
	\begin{itemize}
		\item TAed 15-410 for 3 semesters during undergrad
		\item 1st graduate year was 5th year M.S., rest in Ph.D. program
		\begin{itemize}
			\item \url{http://www.contrib.andrew.cmu.edu/~bblum/thesis.pdf}
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	{\bf About Landslide}
	\begin{itemize}
		\item Simics module, which traces:
			\begin{itemize}
				\item Every instruction executed
				\item Every memory access read/written
			\end{itemize}
			% FIXME for next year -- how much slowdown is there? 20x?
		\item Originally supported only P3s; can now test P2s fully-automated
		\item {\em Landslide} shows how your {\em Pebbles} programs may not be stable.
	\end{itemize}
\end{frame}

@@

\subsection{Technical Overview}

\begin{frame}{Big Picture: Execution Tree Exploration}
	\textbf{Backtracking}
	\begin{itemize}
		\item Each time test completes, identify a PP to replay differently
		\item Reset machine state and start over
		\item Implemented using Simics bookmarks
			\begin{itemize}
				\item {\tt set-bookmark} and {\tt skip-to}
			\end{itemize}
		\item Replay test from the beginning, with a different interleaving
	\end{itemize}
	\pause
	\linegap

	{\bf Controlling scheduling decisions}
	\begin{itemize}
		\item Tool must control all sources of nondeterminism
		\item In 15-410, just timer and keyboard interrupts
			% optional - say: "in real kernels, disk/network io interrupts"
		\item Landslide repeatedly fires timer ticks until desired thread is run.
	\end{itemize}
\end{frame}

\begin{frame}{Landslide \& You}
	\begin{center}
	\includegraphics[width=0.85\textwidth]{landslide-new-blank.pdf}
	\end{center}
\end{frame}

\begin{frame}{Landslide \& You}
	\begin{center}
	\includegraphics[width=0.85\textwidth]{landslide-new.pdf}
	\end{center}
\end{frame}

\begin{frame}{Identifying Bugs}
	Landslide can {\em definitely discover:}
	\begin{itemize}
		\item Assertion failures % Say: note: the better your asserts are..!
		\item Segfaults
		\item Deadlock
		\item Use-after-free / double-free
	\end{itemize}
	\linegap
	Landslide can {\em reasonably suspect:}
	\begin{itemize}
		\item Infinite loop (halting problem)
		\item Data race bugs
	\end{itemize}
\end{frame}

\begin{frame}{What is a Data Race?}
	A {\bf data race} is a pair of memory accesses between two threads, where:
	\begin{itemize}
		\item At least one of the accesses is a write
		\item The threads are not holding the same mutex
		\item The threads can be reordered (e.g., no \texttt{cond\_signal()} in between)
	\end{itemize}
	\pause
	\linegap

	Data races are {\em not necessarily} bugs, just highly suspicious!
	\begin{itemize}
		\item Bakery alg: Is {\tt number[i]=max(number[0],number[1])+1} bad?
		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
		\item What about unprotected {\tt next\_thread\_id++}?
		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
		\begin{itemize}
			\item Hmmm...
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\subsection{Iterative Deepening}

\begin{frame}{Choosing the Right Preemption Points}
	How can we address exponential state space explosion?
	\pause
	\linegap

	State of the art tools hard-code a fixed set of preemption points.
	\begin{itemize}
		\item E.g., ``all thread library API calls'' or ``all kernel mutex locks/unlocks''
		\item Depending on length of test, completion time is unpredictable.
		\item More often, a subset is better in terms of time/coverage. 
	\end{itemize}
	\pause
	\linegap
	Current systematic testing model is not user-friendly.
	\begin{itemize}
		\item Tool: ``I want to use these PPs, but can't predict completion time.''
		\item User: ``I have 16 CPUs and 24 hours to test my program.''
	\end{itemize}
	\linegap
	{\em Stress testing allows user to choose total run time -- can we offer this too?}
\end{frame}

\begin{frame}{Iterative Deepening of Preemption Points}
	Goal: Run the best tests for a given CPU budget.
	\begin{itemize}
		\item Technique: ``Iterative Deepening''
	\end{itemize}
	\linegap

	Based on experience from past 15-410 student volunteers
	\begin{itemize}
		%\item Students worked best with an iterative process
		\item ``Start small, then add more preemption points as time allows''
		\item Landslide now automates this process
	\end{itemize}
	\linegap

@@

\begin{frame}{Iterative Deepening in Landslide}
	Landslide automatically iterates through different configurations of PPs.
	\begin{itemize}
		\item Manages work queue of jobs with different PPs
		\item Each job represents a new state space for Landslide to explore
		\item Prioritizes jobs based on estimated completion time
		%\item Prioritizes which jobs are important / likely to finish
		%\begin{itemize}
		%	\item Based on nature of PPs (data races? mutexes?)
		%	\item Based on estimated completion time
		%\end{itemize}
	\end{itemize}
	\linegap

@@

\begin{frame}{Iterative Deepening}
	Adding different PPs can produce state spaces of different sizes; Landslide tries them in parallel.
	\vspace{0.15in}
	\begin{center}
		\begin{tabular}{cc}
			\includegraphics[width=0.45\textwidth]{tree1.pdf} &
			\includegraphics[width=0.5\textwidth]{tree2.pdf}
		\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{Iterative Deepening}
	If time allows, Landslide will combine PPs into larger, more comprehensive state spaces.
	\begin{center}
		\includegraphics[width=0.8\textwidth]{tree3.pdf}
	\end{center}
	% say: and depending on your time budget, you might decide that you expect this one to be feasible, or you might fall back to a smaller state space from before.
\end{frame}

@@

\breakslide{\Huge Demo}

\begin{frame}{Test Suite}
	Landslide ships with 6 approved test cases:
	\linegap

	Standard P2 tests
	\begin{itemize}
		\item {\tt thr\_exit\_join}
		\item {\tt paraguay}
		\item {\tt rwlock\_downgrade\_read\_test}
	\end{itemize}

	New tests
	\begin{itemize}
		\item {\tt broadcast\_test}
		\item {\tt paradise\_lost} %(new this semester)
		\item {\tt mutex\_test} %(new this semester)
	\end{itemize}
\end{frame}

%\begin{frame}{Preliminary Evaluation}
%	We tested 22 random submitted P2s from Spring 2014.
%	\begin{itemize}
%		\item Test cases:
%		\begin{itemize}
%			\item {\tt thr\_exit\_join}
%			\item {\tt paraguay}
%			\item {\tt rwlock\_downgrade\_read\_test}
%			\item {\tt broadcast\_test} (new)
%		\end{itemize}
%		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
%	\end{itemize}
%\end{frame}
%
%\begin{frame}{Preliminary Results}
%%	88 total tests were run (9,680 total CPU-minutes)
%%	\linegap
%%
%	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
%	\linegap
%
%	56 tests ran to completion without finding bugs:
%	\begin{itemize}
%		\item 47 exhausted the CPU budget.
%		\item 9 completed all state spaces before time ran out.
%	\end{itemize}
%	\linegap
%
%	Between 20 and 33 {\bf non-deterministic} bugs were found:
%	\begin{itemize}
%		\item 20 deadlock races among 12 P2s
%		\item 10 use-after-free races among 6 P2s
%		\item 3 misc (assert fail, segfault) among 2 P2s
%	\end{itemize}
%\end{frame}

\begin{frame}{Last Semester}
	%Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
	%%\begin{itemize}
	%%	\item (that they didn't find during stress testing)
	%%\end{itemize}
	%%%%% hopefully!
	%\linegap

	%Among all groups, 38 total tests were run
	%\begin{itemize}
	%	\item 21 tests ran to completion without finding bugs
	%	\item 4 {\em deterministic} bugs
	%	\item 13 distinct {\em non-deterministic} bugs
	%		\begin{itemize}
	%			\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
	%			\item 4 in {\tt paraguay} (deadlock, infinite loop)
	%			\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
	%			\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
	%		\end{itemize}
	%		%\begin{itemize}
	%		%	\item 6 of which required data-race PPs to expose
	%		%\end{itemize}
	%\end{itemize}
	Fall 2015: 22 groups signed up to use Landslide; 21 found bugs
	% 18 groups found races
	\begin{itemize}
		\item (that they didn't find during stress testing)
	\end{itemize}
	%%%% hopefully!
	\linegap

	Among all groups, 185 distinct tests were run
	\begin{itemize}
		\item 137 tests ran to completion without finding bugs
		\item 21 {\em deterministic} bugs
	\end{itemize}
	\pause % ask: "who can tell me why landslide might find a *deterministic* bug not found during past testing? A: use-after-free
	\linegap

	38 distinct {\em non-deterministic} bugs
	\begin{itemize}
			%\begin{itemize}
			%	\item 7 of which required data-race PPs to expose
			%\end{itemize}
		\item 11 groups (50\%) fixed all their bugs
			\begin{itemize}
				\item (as verified by running Landslide again -- not a guarantee!)
			\end{itemize}
		\item 2.7 bugs ``fixed'' (see above) on average per group
		\item Most ambitious group: 11 distinct races found + fixed!
	\end{itemize}
\end{frame}

%\begin{frame}{A new treat} %since last semester}
%	\texttt{mutex\_test} tries to make your mutexes fail or deadlock.
%	\linegap
%
%	Landslide will look for data races inside your mutex implementation.
%	\begin{itemize}
%		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
%		\item {\tt mutex->whose\_turn\_is\_it = ...;}
%	\end{itemize}
%	\pause
%	\linegap
%	
%	Last semester: 12 non-determinstic bugs among 8 groups
%	\linegap
%
%	%Tested 77 P2s from S'14, F'14, S'15
%	%\begin{itemize}
%	%	\item 5 unrelated deterministic bugs (use-after-free, crash, etc)
%	%		% TODO check these for false positives
%	%		% But this was run after the arbiter fp deadlock avoidance
%	%		% So I think it's ok
%	%	\item 7 deadlock races
%	%	\item 4 other races (NULL crash, infinite loop, use-after-free)
%	%	\item 1 ``two threads got the lock at once'' race
%	%		\begin{itemize}
%	%			\item {\em ...that the TA who graded it didn't find!}
%	%		\end{itemize}
%	%\end{itemize}
%\end{frame}

@@

\subsection{Landslide for the People}

\begin{frame}{User Study}
	\textbf{Try Landslide on your P2!}
	\begin{itemize}
		\item Bare minimum effort: No more than 1 hour
		\begin{itemize}
			\item Clone a github URL, run setup script, run tests %, answer survey questions
			\item Landslide will automatically report test results (as described below)
		\end{itemize}
		\item Full study plan: 4-8 hours of active attention
		\begin{itemize}
			\item (Estimated, including time to diagnose and fix bugs)
			\item However, many tests should run passively overnight -- start soon!
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	Prerequisites
	\begin{itemize}
		\item You {\em must} pass the P2 hurdle before using Landslide.
		\begin{itemize}
			\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
		\end{itemize}
		\item Must have {\em attempted} several stress tests
		\begin{itemize}
			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (15 min), {\tt paraguay}
		\end{itemize}

	\end{itemize}
\end{frame}

\begin{frame}{User Study - Additional Information}
	Human Subjects Research
	\begin{itemize}
		\item CMU IRB has approved this study
		\item Landslide will collect results while you use it
		\begin{itemize}
			\item Record commands issued, take snapshots of your P2 code
			\item All data will be anonymized before publication
		\end{itemize}
		\item No coercion: {\bf \em There is no penalty for not participating.}
		\begin{itemize}
			\item I am not on course staff, cannot influence your grade
			\item Course staff will not have access to study data during semester
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	Risks \& Benefits
	\begin{itemize}
		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
		\item Risk: Landslide may find no bugs and be a waste of your time.
		\item Benefit: You might learn something...
	\end{itemize}
	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
\end{frame}

\begin{frame}{User Study - How to Participate}
	Interested?
	\linegap

	To participate\dots
	\begin{itemize}
		\item Review this lecture and study information sheet
		\item Meet prerequisites of passing P2 tests
		\item Complete sign-up form online to get further instructions
			\begin{itemize}
				\item \url{http://tinyurl.com/landslide-p2-s16}
			\end{itemize}
		\item Optional ``Landslide clinic'' for in-person tech (or moral) support
			\begin{itemize}
				\item Mid-next-week, room and time TBD
			\end{itemize}
		\item Use Landslide in addition to stress tests until P2 is due!
	\end{itemize}
	\linegap

	{\em Really} interested?
	\begin{itemize}
		\item I know of an M.S. thesis topic if you want to do post-OS research.
	\end{itemize}
\end{frame}

>>>>

\title[Landslide]{{\bf Stateless Model Checking \\ with Data-Race Preemption Points}}
%\subtitle[]{CALCM group seminar}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[]{Carnegie Mellon University - CALCM group seminar}
\date[]{2016, March 01}

@@

\begin{frame}{}
	Working on OOPSLA submission (March 23)
	\linegap

	Feedback wanted on:
	\begin{itemize}
		\item What do OOPSLA reviewers look for?
		\item Collaboration and/or ideas for future work
	\end{itemize}
\end{frame}

\begin{frame}{Outline}
	\textbf{Background}
	\begin{itemize}
		\item Preemption points
		\item Stateless model checking (aka, systematic testing)
		\item Data races
	\end{itemize}
	{\bf Quicksand \& Iterative Deepening}
	\begin{itemize}
		\item State space estimation
		\item Data-race preemption points
		\item Implementation with Landslide
	\end{itemize}
	{\bf Evaluation}
	\begin{itemize}
		\item Comparison to ``single-state-space'' approach
		\item Nondeterministic data-race candidates
		\item Providing total verification
	\end{itemize}
\end{frame}

\section{Introduction}

\subsection{Stateless Model Checking}
\definecolor{thread1}{RGB}{87,172,255}
\definecolor{thread2}{RGB}{255,201,102}
\definecolor{thread3}{RGB}{255,128,160}

@@

		\begin{itemize}
			%\item See {\bf Paradise Lost} lecture!
			\item {\tt if} vs {\tt while}: Two consumers can interleave to make one fail.
		\end{itemize}
	\end{center}
\end{frame}
\begin{frame}{Thread Interleavings (``good'' case)}
	\begin{center}
		\begin{tabular}{|l|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
			\hline
			\small \texttt{lock(mx);} & & \\
			\small \texttt{if~(!work\_exists())} & & \\
			\small \texttt{~~~~wait(cvar,~mx);} & & \\

@@

\newcommand\related[1]{\textsuperscript{\em [#1]}}

\begin{frame}{Motivating Stateless Model Checking}
	Goal: Test as many interleavings as you can (given some CPU budget)
	\linegap

	\textbf{Stress testing}
	\begin{itemize}
		\item Relies on randomly-occurring timer (or I/O) interrupts
			% to perturb threads
		\item Exposes nondeterministic bugs at random
		\item Might not find bugs in any bounded time
	\end{itemize}
	\linegap

	\textbf{Stateless Model Checking} (MC)\related{Godefroid '97}
	\begin{itemize}
		\item Exhaustively exercise {\em all} thread interleavings
		\begin{itemize}
			\item ...around a given set of {\em preemption points}
		\end{itemize}
		\item Identify bugs for certain (e.g. segfault, deadlock, panic)
		\item More reliable coverage, better reproducibility
		\item Exponentially-sized state space of interleavings
			%depending on \# of preemption points
	\end{itemize}
	%Moving beyond stress testing: which {\em preemption points} are important?
\end{frame}

\begin{frame}{Preemption Points}
	\textbf{Preemption points} (PPs) are program locations where a context switch might expose different behaviour.
	\linegap

	Stateless MC is {\em parameterized} by which PPs can trigger thread switches.
	\linegap

	For case study, {\tt mutex\_lock}, {\tt mutex\_unlock}, {\tt cond\_wait} are relevant.
	\pause
	\linegap

	Which PPs to use?
	\begin{itemize}
		\item Every instruction (exhaustive, but intractable)
		\item Synchronization API calls\related{Musuvathi '08, Simsa '10, Blum '12}
		%\item Kernel mutex interface\related{Blum '12}
		\item Global/heap memory accesses\related{Holzmann '97, Yang '08}
	\end{itemize}
\end{frame}

\begin{frame}{MC State of the Art}
	Modern stateless MC tools can {\em mitigate} exponential explosion.
	\begin{itemize}
		\item Dynamic Partial Order Reduction\related{Flanagan '05}
		\item State space estimation\related{Simsa '12}
	\end{itemize}
	\linegap

	However, existing MCs ``hard-code'' their PPs in advance.
	\begin{itemize}
		\item With too many preemption points, tests time out.
		\item Which PPs are ``good'' differs for each test case.
		\item Relationship between PPs and testing time is unpredictable.
	\end{itemize}
	\linegap

	Current MC model is not user-friendly.
	\begin{itemize}
		\item Tool: ``I want to use these PPs, but can't predict completion time.''
		\item User: ``I have 16 CPUs and 24 hours to test my program.''
	\end{itemize}
\end{frame}

@@

\subsection{Data Races}

\begin{frame}{Data Races\related{Savage '97}}
	A {\bf data race} is a pair of memory accesses between two threads, where:
	\begin{itemize}
		\item At least one of the accesses is a write
		\item The threads are not holding the same mutex
		%\item The threads can be reordered (e.g., no \texttt{cond\_signal()} in between)
		\item The threads' executions are not related by {\em Happens-Before}
	\end{itemize}

	Data races are {\em not necessarily} bugs, just highly suspicious!
	\begin{itemize}
		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
		\item Hence, data race locations are ideal PP candidates.
	\end{itemize}
\end{frame}

\begin{frame}{The Happens-Before (HB) Relation}
	Most data-race analyses use {\em Pure HB} to determine simultaneity.
	\begin{itemize}
		\item HB is established by any synchronization (mutexes, cvars, xchg...)
		\item Represents possiblity to threads' behaviour to change if reordered
			%\begin{itemize}
			%	\item ...that the programmers obviously intended.
			%\end{itemize}
		\item May exclude some races from alternate interleavings ({\em false negatives})
	\end{itemize}
	\pause
	\linegap

	To avoid false negatives, we use {\em Limited HB}\related{O'Callahan '03} instead.
	%, Serebryany '09}
	\begin{itemize}
		\item HB is established only when thread runnability changes (cvars).
		\item Represents possibility to reorder accesses to occur simultaneously
		\item Captures accesses separated by unrelated, synchronized, logic
		\item Will also falsely flag some {\em false positives}
		\item Combined with lock-set analysis to increase accuracy
	\end{itemize}
\end{frame}

\begin{frame}{Happens-Before Example}
	\begin{center}
		\begin{tabular}{|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
			\hline
			& \\
			\small \texttt{x++;} // A & \\
			\small \texttt{lock(mx);} & \\
			%\small \texttt{y = true;} & \\
			\small \texttt{unlock(mx);} & \\

			& \\
			& \small \texttt{lock(mx);} \\
			& \small \texttt{unlock(mx);} \\
			& \small \texttt{x++;} // B \\
			& \\
			\hline
		\end{tabular}
	\end{center}
	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (false negative).
	\linegap

	Limited HB: $\neg HB(A, B)$; data race is reported (correctly).
\end{frame}

\begin{frame}{Happens-Before Example}
	\begin{center}
		\begin{tabular}{|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
			\hline
			\small \texttt{x++;} // A & \\
			\small \texttt{lock(mx);} & \\
			\small \texttt{y = true;} & \\
			\small \texttt{unlock(mx);} & \\

			& \small \texttt{lock(mx);} \\
			& \small \texttt{bool z = y;} \\
			& \small \texttt{unlock(mx);} \\
			& \small \texttt{if (z);} \\
			& \small \texttt{~~x++;} // B \\
			\hline
		\end{tabular}
	\end{center}
	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (correctly).
	\linegap

	Limited HB: $\neg HB(A, B)$; data race is reported (false positive).
\end{frame}

@@

\breakslide{\Huge Quicksand}

\begin{frame}{Iterative Deepening of Preemption Points}
	Goal: Run the best tests for a given CPU budget.
	\linegap

	Inspired by past experience with Landslide, our MC tool\related{Blum '12}
	\begin{itemize}
		\item 15-410 student volunteers worked best with iterative process
		\item ``Start small, then add more preemption points as time allows''
		\item Quicksand now automates this via {\em Iterative Deepening}
	\end{itemize}
	\linegap

@@

\begin{frame}{Iterative Deepening in Quicksand}
	Quicksand automatically iterates through different configurations of PPs.
	\begin{itemize}
		\item Manages work queue of jobs with different PPs
		\item Each job represents a new state space for Landslide to explore
		\item Prioritizes jobs based on estimated completion time
		%\item Prioritizes which jobs are important / likely to finish
		%\begin{itemize}
		%	\item Based on nature of PPs (data races? mutexes?)
		%	\item Based on estimated completion time
		%\end{itemize}
		\item New PPs added on-the-fly from data-race analysis
	\end{itemize}
	\linegap

@@

\begin{frame}{Iterative Deepening}
	Adding different PPs can produce state spaces of different sizes; Quicksand tries them in parallel, as informed by state space estimation\related{Simsa '12}.
	\vspace{0.15in}
	\begin{center}
		\begin{tabular}{cc}
			\includegraphics[width=0.45\textwidth]{tree1.pdf} &
			\includegraphics[width=0.5\textwidth]{tree2.pdf}
		\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{Iterative Deepening}
	If time allows, combine PPs into larger, more comprehensive state spaces.
	\begin{center}
		\includegraphics[width=0.8\textwidth]{tree3.pdf}
	\end{center}
	% say: and depending on your time budget, you might decide that you expect this one to be feasible, or you might fall back to a smaller state space from before.
\end{frame}

\begin{frame}{Total Verification}
	% Let's come back to this issue of ...
	MC can {\em totally verify} a test by testing a state space with all possible PPs.
	\begin{itemize}
		\item 
	\end{itemize}
	%One downside of hard-coding PPs in advance is timing out when the test is too large, but another downside is if the test is small enough, you might miss a possible full verification!
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@
